Whittaker Models for Real Groups

### 2024-04-02 whittaker.tex
Small typos

Suppose $\pi$ is an irreducible representation of~$G(\R)$. Recall $\pi$ \emph{has a Whittaker model} when there exists a pair $(B(\R), \chi)$ where 
\begin{enumerate}[(i)]
\item $B$ is a Borel subgroup of~$G$ defined over~$\R$, with real points $B(\R)$;
\item $\eta$ is a character of the nilpotent radical~$N(\R)$ of~$B(\R)$, subject to the condition that~$\eta$ is nontrivial on each simple root space;
\end{enumerate}
such that~$\pi$ embeds into the induced representation $\mathrm{Ind}_{N(\R)}^{G(\R)}(\chi)$. Of course this can happen if and only if there exists a Borel subgroup $B \subset G$ satisfying~(i); by definition this means~$G$ is \emph{quasisplit}, which we assume from now on.

@@

We may view $\mathfrak{w}(\pi)$ as a useful invariant of generic representations. For instance, suppose~$\pi$ is a discrete series representation of~$G(\R)$. Then $\mathfrak{w}(\pi)$ singles out~$\pi$ among the discrete series representations of~$G(\R)$ having  the same infinitesimal and central characters. In the language of the local Langlands correspondence this means the Whittaker datum singles out~$\pi$ within its $L$-packet.

More generally, Whittaker data play a subtle but key role in the local Langlands correspondence for reductive groups over arbitrary local fields (including the nonarchimedean case). For instance, given a Whittaker datum~$\mathfrak{w}$, it is expected that every tempered $L$-packet contains precisely one $\mathfrak{w}$-generic representation, which can be used as a basepoint for the internal structure of the $L$-packet. (Here insert refs to the work of Tasho and others.)

@@

The following uses rather deep results of representation theory:
\begin{proposition} Let~$\pi$ be the $(\g, K)$-module for a large discrete series representation of~$G(\R)$.  
\begin{enumerate}[(a)]
\item There is a unique Whittaker datum~$\mathfrak{w}$ such that~$\pi$ is $\mathfrak{w}$-generic.
\item $\AV(\pi)$ is the closure of a single $K$-orbit on~$\mathcal{O}_p \cap \mathfrak{s}$.
\item  $\WF(\pi)$ is the closure of a single $G(\R)$-orbit on $\mathcal{O}_p \cap  i \g(\R)$.
\end{enumerate}
\end{proposition}

@@

Given a large discrete series representation~$\pi$ we consider the invariants of~$\pi$ defined by the Proposition: a Whittaker datum  $\mathfrak{w}(\pi)$, a $K$-orbit $\AV^\flat(\pi) \in (\mathcal{O}_p \cap \mathfrak{s})/K$  and a nilpotent $G(\R)$-orbit $\WF^\flat(\pi) \in (\mathcal{O}_p \cap  i \g(\R))/G(\R)$.

@@

Using the group cohomology interpretation in the previous paragraph, we can construct a canonical  isomorphism between $Q_{\sigma}(G)$ and $Q_{\theta}(G)$.
For this recall $\theta\sigma = \sigma\theta$ and $G^{\sigma\theta}$ is a compact real form of~$G$.
Furthermore $\sigma$ and $\theta$ coincide on $G^{\sigma\theta}$,
therefore  $H_{\sigma}(\Gamma, G^{\sigma\theta})$ and  $H_{\sigma}(\Gamma, G^{\sigma\theta})$ are the same set,
which we will denote by  $H_{\sigma\mid\theta}(\Gamma, G^{\sigma\theta})$.
Now the inclusion $G^{\sigma\theta} \hookrightarrow G$ induces canonical maps 
\begin{equation} \label{bij_cohom} H^{1}_{\sigma}(\Gamma, G) \longleftarrow H_{\sigma\mid\theta}(\Gamma, G^{\sigma\theta}) \longrightarrow H^{1}_{\theta}(\Gamma, G).\end{equation}
By \cite[Corollary~4.4 and Corollary 4.7]{galois}, both of these maps are \emph{bijections}.
Replacing~$G$ by $Z$ we get bijections $H^{1}_{\sigma}(\Gamma, Z) \leftarrow H_{\sigma\mid\theta}(\Gamma, Z^{\sigma\theta}) \rightarrow H^{1}_{\theta}(\Gamma, Z)$.
These fit with the bijections~\eqref{bij_cohom} into a commutative diagram
\[
\begin{CD}
H^{1}_{\sigma}(\Gamma, Z) @<<< H_{\sigma\mid\theta}(\Gamma, Z^{\sigma\theta})@>>> H^{1}_{\theta}(\Gamma, Z)
 \\
@V{\varphi_{\sigma}}VV @VVV @V{\varphi_{\theta}}VV  \\
H^{1}_{\sigma}(\Gamma, G) @<<< H_{\sigma\mid\theta}(\Gamma, G^{\sigma\theta})@>>> H^{1}_{\theta}(\Gamma, G)
\end{CD}
\]
where $\varphi_{\sigma}$ and $\varphi_{\theta}$ are induced by the inclusion $Z \hookrightarrow G$. Since the horizontal arrows are isomorphisms in the category of pointed sets, they induce a canonical bijection between the kernels of $\varphi_{\sigma}$ and $\varphi_{\theta}$; by~\eqref{interp_q_cohomology} this gives a canonical bijection 
\[ Q_{\sigma}(G) \simeq Q_{\theta}(G).\] 
As discussed after~\eqref{interp_q_cohomology}, this is actually an isomorphism of abelian groups.

>>>>

Suppose $\pi$ is an irreducible representation of~$G(\R)$. Recall $\pi$ has a \emph{Whittaker model} when there exists a pair $(B(\R), \chi)$ where 
\begin{enumerate}[(i)]
\item $B$ is a Borel subgroup of~$G$ defined over~$\R$, with real points $B(\R)$;
\item $\eta$ is a character of the nilpotent radical~$N(\R)$ of~$B(\R)$, subject to the condition that~$\eta$ is nontrivial on each simple root space;
\end{enumerate}
such that~$\pi$ embeds into the induced representation $\mathrm{Ind}_{N(\R)}^{G(\R)}(\chi)$. Of course this can happen if and only if there exists a Borel subgroup $B \subset G$ satisfying~(i); by definition this means~$G$ is \emph{quasisplit}, which we assume from now on.

@@

We may view $\mathfrak{w}(\pi)$ as a useful invariant of generic representations. For instance, suppose~$\pi$ is a discrete series representation of~$G(\R)$. Then $\mathfrak{w}(\pi)$ singles out~$\pi$ among the discrete series representations of~$G(\R)$ having  the same infinitesimal and central characters. In the language of the local Langlands correspondence \cite{Langlands_CIRRAC, Contragredient}, this means the Whittaker datum singles out~$\pi$ within its $L$-packet.

More generally, Whittaker data play a subtle but key role in the local Langlands correspondence for reductive groups over arbitrary local fields \cite{Borel, Kaletha_Taibi}, including the nonarchimedean case. For instance, given a Whittaker datum~$\mathfrak{w}$, it is expected that every tempered $L$-packet contains precisely one $\mathfrak{w}$-generic representation, which can be used as a basepoint for the internal structure of the $L$-packet~\cite{Kaletha_genericity}.

@@

\subsection{The invariants for large discrete series}

The following uses rather deep results of representation theory.

\begin{proposition} \label{invariants_ds}
Let~$\pi$ be the $(\g, K)$-module for a large discrete series representation of~$G(\R)$.  
\begin{enumerate}[(a)]
\item There is a unique Whittaker datum~$\mathfrak{w}$ such that~$\pi$ is $\mathfrak{w}$-generic.
\item $\AV(\pi)$ is the closure of a single $K$-orbit on~$\mathcal{O}_p \cap \mathfrak{s}$.
\item  $\WF(\pi)$ is the closure of a single $G(\R)$-orbit on $\mathcal{O}_p \cap  i \g(\R)$.
\end{enumerate}
\end{proposition}

@@

Given a large discrete series representation~$\pi$ we consider the invariants of~$\pi$ defined by Proposition~\ref{invariants_ds}: a Whittaker datum  $\mathfrak{w}(\pi)$, a $K$-orbit $\AV^\flat(\pi)$ in $\mathcal{O}_p \cap \mathfrak{s}$, and a $G(\R)$-orbit $\WF^\flat(\pi)$ in $(\mathcal{O}_p \cap  i \g(\R))$.

@@

Using the group cohomology interpretation in the previous paragraph, we can construct a canonical  isomorphism between $Q_{\sigma}(G)$ and $Q_{\theta}(G)$.
For this recall $\theta\sigma = \sigma\theta$ and $G^{\sigma\theta}$ is a compact real form of~$G$.
Furthermore $\sigma$ and $\theta$ coincide on $G^{\sigma\theta}$,
therefore  $H_{\sigma}(\Gamma, G^{\sigma\theta})$ and  $H_{\theta}(\Gamma, G^{\sigma\theta})$ are the same set,
which we will denote by  $H_{\sigma\mid\theta}(\Gamma, G^{\sigma\theta})$.
Now the inclusion $G^{\sigma\theta} \hookrightarrow G$ induces canonical maps 
\begin{equation} \label{bij_cohom} H^{1}_{\sigma}(\Gamma, G) \longleftarrow H_{\sigma\mid\theta}(\Gamma, G^{\sigma\theta}) \longrightarrow H^{1}_{\theta}(\Gamma, G).\end{equation}
By \cite[Corollary~4.4 and Corollary 4.7]{galois}, both of these maps are \emph{bijections}.
Replacing~$G$ by $Z$ we get bijections $H^{1}_{\sigma}(\Gamma, Z) \leftarrow H_{\sigma\mid\theta}(\Gamma, Z^{\sigma\theta}) \rightarrow H^{1}_{\theta}(\Gamma, Z)$.
These fit with the bijections~\eqref{bij_cohom} into a commutative diagram
\[
\begin{CD}
H^{1}_{\sigma}(\Gamma, Z) @<<< H_{\sigma\mid\theta}(\Gamma, Z^{\sigma\theta})@>>> H^{1}_{\theta}(\Gamma, Z)
 \\
@V{\varphi_{\sigma}}VV @VVV @V{\varphi_{\theta}}VV  \\
H^{1}_{\sigma}(\Gamma, G) @<<< H_{\sigma\mid\theta}(\Gamma, G^{\sigma\theta})@>>> H^{1}_{\theta}(\Gamma, G)
\end{CD}
\]
where $\varphi_{\sigma}$ and $\varphi_{\theta}$ are induced by the inclusion $Z \hookrightarrow G$. Since the horizontal arrows are isomorphisms in the category of pointed sets, they induce a canonical bijection between the kernels of $\varphi_{\sigma}$ and $\varphi_{\theta}$; by~\eqref{interp_q_cohomology} this gives a canonical bijection 
\[ Q_{\sigma}(G) \simeq Q_{\theta}(G).\] 
As discussed after~\eqref{interp_q_cohomology}, this is actually an isomorphism of abelian groups.
